Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
done up to VII

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precisely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quantifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$.

@@

\begin{proof}
  We only show the cases for $\wfaD$ and $\cfaD$:
  $$
  \todo{}
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\fequ}{(x \notin fv(A))}{\forall x.(A \vlor A)}{
        \vlin{\wrD}{}{A \vlor (\forall x.A)}{
          \vlhy{A}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  $$
  The other cases have been shown before (e.g.~\cite{brunnler:phd}).
%%   We have the following derivations:
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.A}{
%%       \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
%%         \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
%%           \vlhy{\forall x.\forall x.A}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
%%       \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
%%       \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
%% 	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
%% 	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
%% 	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
%% 	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\exists x.(A \vlor B)}{
%%       \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
%% 	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.(A \vlor B)}{
%%       \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
%% 	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
%%   $$.
%%   \juihsuan{done. If needed, we can introduce the notion of open deduction to
%% reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

\begin{lemma}
  Let $A$ and $B$ be formulas. Then
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{B}{
        \vlhy{A}}}
    \qquad
    \iff
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}
  \end{equation*}
\end{lemma}

@@

\begin{proof}
  First, we can replace every instance of the $\fff$-rule by
  $\weakr$. Then the instances of $\weakr$ and $\conr$ are replaced by
  $\wD$ and $\cD$, which can then be permuted down. (Details are in
  Appendix~\ref{app:LK1-decompose}.
\end{proof}

@@

\begin{proof}
  First, observe that the $\exists$ rule can be permuted downwards over all the
other rules since $A\ssubst{x}{t}$ has the same structure as $A$ and none of the
other rules has a premise of the form $\Scons{\exists x.A}$. It suffices now to
prove that for all $r_1 \in \set{\forall, \aiD, \tttD}$, for all $r_2 \in
\set{\sw, \mix, \fequ}$, we can permute $r_1$ upwards over $r_2$.
  We show some cases here, and leave the others to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\aiD}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\sw}{}{\Scons{(A \vlan \ttt) \vlor C}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\sw}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\aiD}{}{\Scons{A \vlan ((a \vlor \cneg a) \vlor C)}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \end{equation*}
  \begin{equation*}
    \vlderivation{
      \vlin{\tttD}{}{\Scons{(A \vlor (B \vlan \ttt))}}{
        \vlin{\mix}{}{\Scons{A \vlor B}}{
          \vlhy{\Scons{A \vlan B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mix}{}{\Scons{(A \vlor (B \vlan \ttt)}}{
      \vlin{\tttD}{}{\Scons{A \vlan (B \vlan \ttt)}}{
        \vlhy{\Scons{A \vlan B}}}}}
  \end{equation*}

  \tocheck{}
\end{proof}

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}

\begin{proof}

  \todo{permutation with $\fequ$}

@@

\begin{proof}[Proof of Theorem~\ref{thm:decomposition}]
  Assume we have a proof of $A$ in $\FOKS$. We first apply
  Theorem~\ref{thm:decompositionA}, and then
  Lemma~\ref{lem:MLS1-decomposition} to the upper half and
  Lemma~\ref{lem:cw-decomposition} to the lower half.
\end{proof}

@@

\subsection{Rule permutation for the proof of Lemma~\ref{lem:cw-decomposition}}

>>>>

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precisely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quantifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$. To give an example, consider
the derivation in Figure~\ref{fig:example-decompose} which corresponds
to the right-most combinatorial proof in Figures~\ref{fig:cps}
and~\ref{fig:cps-condensed}.

@@

\begin{proof}
  We only show the cases for $\wfaD$ and $\cfaD$ (for the others see~\cite{brunnler:phd}):
  \vadjust{\vskip-2ex}
  \begin{equation}
    \label{eq:wfa-cfa}
  \qquad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\fequ}{}{\forall x.(A \vlor A)}{
        \vlin{\wrD}{}{A \vlor (\forall x.A)}{
          \vlhy{A}}}}}
  \qquad
  \quad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \end{equation}
  where in the first derivation, $x$ is not free in $A$, and in the second one not free in $\forall x.A$.
%  The other cases have been shown before (e.g.~\cite{brunnler:phd}).
%%   We have the following derivations:
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.A}{
%%       \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
%%         \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
%%           \vlhy{\forall x.\forall x.A}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
%%       \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
%%       \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
%% 	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
%% 	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
%% 	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
%% 	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\exists x.(A \vlor B)}{
%%       \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
%% 	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.(A \vlor B)}{
%%       \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
%% 	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
%%   $$.
%%   \juihsuan{done. If needed, we can introduce the notion of open deduction to
%% reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

\begin{lemma}\label{lem:cw-atomic}
  Let $A$ and $B$ be formulas. Then
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{B}{
        \vlhy{A}}}
    \qquad
    \iff
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}
  \end{equation*}
\end{lemma}

@@

\begin{proof}
  First, we can replace every instance of the $\fff$-rule in $\Deri$ by
  $\weakr$. Then the instances of $\weakr$ and $\conr$ are replaced by
  $\wD$ and $\cD$, which can then be permuted down. (Details are in
  Appendix~\ref{app:LK1-decompose}.
\end{proof}

@@

\begin{proof}
  First, observe that the $\exists$ rule can be permuted downwards
  over all the other rules since $A\ssubst{x}{t}$ has the same
  structure as $A$ and none of the other rules has a premise of the
  form $\Scons{\exists x.A}$. It suffices now to prove that all rules
  in $\set{\forall, \aiD, \tttD}$ can be permuted over the rules in
  $\set{\sw, \mix, \fequ}$, which is straightforward
  (see~\cite{dissvonlutz} for details).
\end{proof}
%% we can permute $r_1$ upwards over $r_2$.
%%   We show some cases here, and leave the others to the reader.
%%   \begin{equation*}
%%   \vlderivation{
%%     \vlin{\aiD}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
%%       \vlin{\sw}{}{\Scons{(A \vlan \ttt) \vlor C}}{
%%         \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
%%   \leadsto
%%   \vlderivation{
%%     \vlin{\sw}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
%%       \vlin{\aiD}{}{\Scons{A \vlan ((a \vlor \cneg a) \vlor C)}}{
%%         \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
%%   \end{equation*}
%%   \begin{equation*}
%%     \vlderivation{
%%       \vlin{\tttD}{}{\Scons{(A \vlor (B \vlan \ttt))}}{
%%         \vlin{\mix}{}{\Scons{A \vlor B}}{
%%           \vlhy{\Scons{A \vlan B}}}}}
%%   \leadsto
%%   \vlderivation{
%%     \vlin{\mix}{}{\Scons{(A \vlor (B \vlan \ttt)}}{
%%       \vlin{\tttD}{}{\Scons{A \vlan (B \vlan \ttt)}}{
%%         \vlhy{\Scons{A \vlan B}}}}}
%%   \end{equation*}

%%   \tocheck{}
%% \end{proof}

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}

\begin{proof}
  We first permute all instances of $\wD$ and $\wfaD$ to the bottom of
  the derivation and then permute in a second step the rules $\cD$ and
  $\cfaD$ below $\set{\me,\mfaD,\mexD}$. This involves a tedious but
  straightforward case analysis. However, unlike most other rule
  permutations in this paper this has not been done before in the deep
  inference literature. For this reason, we give the full case
  analysis in Appendix~\ref{app:cw-decomposition}. Note that this
  Lemma is the reason for the presence of the rules $\wfaD$ and
  $\cfaD$, as without them the permutation cases in~\eqref{eq:wfa-cfa}
  could not be resolved.
\end{proof}

@@

\begin{proof}[Proof of Theorem~\ref{thm:decomposition}]
  Assume we have a proof of $A$ in $\FOKS$. We first apply
  Theorem~\ref{thm:decompositionA}, and then
  Lemma~\ref{lem:MLS1-decomposition} to the upper half and
  Lemmas~\ref{lem:cw-atomic} and~\ref{lem:cw-decomposition} to the lower half.
\end{proof}

@@

\subsection{Rule permutation for the proof of Lemma~\ref{lem:cw-decomposition}}
\label{app:cw-decomposition}
